---
title: "Compare bulkRNA and protein data"
author: "Kaspar Bresser"
date: "14/08/2025"
output: 
  github_document:
    toc: true
  html_document: 
    theme: simplex
    highlight: pygments
#    code_folding: show
    self_contained: TRUE
    toc: yes
    toc_float:
      collapsed: no
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
                      fig.width = 5,
                      fig.asp = 0.618,
                      fig.align = "center")
```



```{r loading, echo=FALSE}
library(tidyverse)
library(lemon)
library(readxl)
library(limma)
library(ggrastr)
library(gghighlight)
library(rstatix)
```


Make some example vizualisations 

##

Import data


```{r}
dat.combined <- read_tsv( "Output/data_combined.tsv")
dat.DE.all <- read_tsv("Output/data_DE_all.tsv")

```



## Selected proteins

Plot proteins that were selected for follow-up


```{r fig.width=11, fig.asp=.35}
prots <-  c("FASN", "CHD4", "DDX3X", "TPP2", "HMGN4", "PTGES2",  "CRYZ", "APEX1", "PRCP",
            "ACAT2", "ACLY", "CAPRIN1", "PTPN22", "STAT3")

dat.combined %>% 
  dplyr::filter(gene %in% prots) %>% 
  mutate(gene = factor(gene, levels = prots), count = log2(count)) %>% 
  pivot_longer(cols = c(abundance, count), names_to = "metric", values_to = "value") %>% 
ggplot(aes(x = condition, y = value))+
  geom_boxplot(aes(fill = condition))+
  geom_jitter(aes(fill = condition), shape = 21, size = 2, width = .1)+
  scale_fill_manual(values = rev(c("#cb181d", "#fc9272", "lightblue")))+
  facet_rep_wrap(metric~gene, scales = "free", nrow = 2)+
  theme_classic()+
  theme(panel.grid.major = element_line(color = "grey90"), legend.title = element_blank(), 
        strip.background = element_blank(), legend.position = "bottom", axis.text.x = element_blank())

ggsave("Figs/MSvRNA_boxplots_selection.pdf", width = 12, height = 4)
```

Specifically plot tumor-derived populations

```{r fig.width=11, fig.asp=.35}
dat.combined %>% 
  filter(str_detect(condition, "Tumor")) %>% 
  dplyr::filter(gene %in% prots) %>% 
  mutate(gene = factor(gene, levels = prots), count = log2(count)) %>% 
  pivot_longer(cols = c(abundance, count), names_to = "metric", values_to = "value") %>% 
ggplot(aes(x = condition, y = value))+
  geom_boxplot(aes(fill = condition))+
  geom_jitter(aes(fill = condition), shape = 21, size = 2, width = .1)+
  scale_fill_manual(values = rev(c("#cb181d", "#fc9272")))+
  facet_rep_wrap(metric~gene, scales = "free", nrow = 2)+
  theme_classic()+
  theme(panel.grid.major = element_line(color = "grey90"), legend.title = element_blank(), 
        strip.background = element_blank(), legend.position = "bottom", axis.text.x = element_blank())

#ggsave("Figs/dotplots_correlation.pdf", width = 3.8, height = 2.7)
```


## Correlations

Calculate Spearman correlation between protein abundance and gene counts

```{r}
dat.combined %>% 
  group_by(gene) %>%
  mutate(count = log2(count)) %>% 
  cor_test(count, abundance, method = "spearman") -> dat.cors
```

Combine with DE data

```{r}
dat.cors %>% 
  dplyr::select(gene, cor, p) %>% 
  inner_join(dat.DE.all, by = c("gene" = "gene.symbol")) -> dat.all.stats
```


```{r}
dat.all.stats %>% 
  filter(P.Value.prot < 0.05 & p > 0.05)
```


Output stats 

```{r}
write_tsv(dat.all.stats, "Output/data_DEandCors.tsv")
```




Check out some interesting proteins

```{r fig.width=7, fig.asp=.55}
prots <- c("ANXA1",  "ANXA2", "SP100", "MYO1E", "GZMA", "FASN",
           "NFKB2", "PDLIM5", "MSH6", "PCNA", "GZMK", "TALDO1")

dat.combined %>% 
  dplyr::filter(gene %in% prots) %>% 
  mutate(gene = factor(gene, levels = prots)) %>% 
  ggplot(aes(x = log2(count), y = abundance))+
  geom_point(aes(fill = condition), shape = 21, size = 2)+
  scale_fill_manual(values = rev(c("#cb181d", "#fc9272", "lightblue")))+
  facet_rep_wrap(~gene, scales = "free", nrow = 2)+
  theme_classic()+
  theme(panel.grid.major = element_line(color = "grey92"), legend.title = element_blank(), strip.background = element_blank(), legend.position = "bottom")+
  labs(x = "RNA expression level", y = "protein expression level")

ggsave("Figs/dotplots_correlation.pdf", width = 7, height = 3.5)
```

Make dot plots for selected genes to show correlation 


```{r fig.width=14, fig.asp=.18}
prots <- c("GZMA",  "GZMK", "AHNAK", "NFATC2", "STAT4", 
           "NFKB2", "SUZ12",  "LYZ", "CD48")

p1 <- dat.combined %>% 
  dplyr::filter(gene %in% prots) %>% 
  mutate(name = factor(gene, levels = prots)) %>% 
ggplot(aes(x = log2(count), y = abundance))+
  geom_point(aes(fill = condition), shape = 21, size = 2)+
  scale_fill_manual(values = rev(c("#cb181d", "#fc9272", "lightblue")))+
  facet_rep_wrap(~name, scales = "free", nrow = 1)+
  theme_classic()+
  theme(panel.grid.major = element_line(color = "grey92"), legend.title = element_blank(), strip.background = element_blank(), legend.position = "bottom")+
  labs(x = "RNA expression level", y = "protein expression level")
```


Also make plots as boxes


```{r fig.width=14, fig.asp=.35}


p2 <- dat.combined %>% 
  dplyr::filter(gene %in% prots) %>% 
  mutate(gene = factor(gene, levels = prots), count = log2(count)) %>% 
  pivot_longer(cols = c(abundance, count), names_to = "metric", values_to = "value") %>% 
ggplot(aes(x = condition, y = value))+
  geom_boxplot(aes(fill = condition))+
  geom_jitter(aes(fill = condition), shape = 21, size = 2, width = .1)+
  scale_fill_manual(values = rev(c("#cb181d", "#fc9272", "lightblue")))+
  facet_rep_wrap(metric~gene, scales = "free", ncol = length(prots))+
  theme_classic()+
  theme(panel.grid.major.y = element_line(color = "grey90"), legend.title = element_blank(), 
        strip.background = element_blank(), legend.position = "bottom", axis.text.x = element_blank())
```


Combine and visualize


```{r fig.width=12}
library(ggpubr)

ggarrange(plotlist = list(p2, p1), nrow = 2, heights = c(2.15,1))


ggsave("Figs/example_plots.pdf", width = 9.5, height = 6.5)
```










