---
title: "scRNAseq NSCLC cyt/dys scores"
author: "Kaspar Bresser"
date: "18/012/2025"
output: 
  github_document:
    toc: true
  html_document: 
    theme: simplex
    highlight: pygments
#    code_folding: show
    self_contained: TRUE
    toc: yes
    toc_float:
      collapsed: no
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
                      fig.width = 5,
                      fig.asp = 0.618,
                      fig.align = "center")
```

Comparative analysis of the cytotoxicity/dysfunction gene-scores to the CITE-seq antibody stainings

```{r loading, echo=TRUE}
library(Seurat)
library(metacell)
library(tidyverse)
library(lemon)
library(ggpubr)
library(rstatix)
```


## Import data

Import the MetaCell object

```{r import_mc}
scdb_init("Data")
mc <- scdb_mc("NSCLC_MC_reordered")

```


Extract the patient info from the mat object

```{r get_hashtags}
mat.obj <- scdb_mat("NSCLC_filt")

mat.obj@cell_metadata %>% 
  select(Patient) %>% 
  as_tibble(rownames = "cellcode") -> patients

patients
```


## Gene expression 

Import Seurat object and Normalize the data


```{r}
seurat.obj <- read_rds("Data/CD8_NSCLC_filt.rds")
seurat.obj <- NormalizeData(seurat.obj, assay = "RNA", normalization.method = "CLR")
```

Define function to extract counts per gene per cellcode from the seurat object


```{r}
get_expression <- function(gene.list){
  
 seurat.obj %>% 
  GetAssayData( layer = "data", assay = "RNA")%>% 
  t() %>% 
  as.data.frame() %>% 
  as_tibble(rownames = "cellcode") %>% 
  select(one_of(c("cellcode", gene.list)))
  
}
```




## Check scores by surface expression


get the CITEseq information

```{r}
mat.obj@cell_metadata %>% 
  as_tibble(rownames = "cellcode") %>% 
  select(cellcode, contains("UMIA")) -> CITE

CITE
```

Normalize the CITEseq counts between patients and take the median of each CITEseq antibody per MetaCell

```{r}
MC.ids <- enframe(mc@mc, name = "cellcode", value = "MetaCell") %>% mutate(MetaCell = as.character(MetaCell))

CITE %>% 
  inner_join(MC.ids) %>% 
  inner_join(patients) %>% 
  group_by(Patient) %>% 
  mutate(across(contains("UMIA"), ~(((.+1)/sum(.))*10000))) %>% 
  group_by(MetaCell, Patient) %>% 
  summarise(across(contains("UMIA"), median),
            n = n()) -> average.CITE

average.CITE
```


Visualize per patient, and set some cutoffs (manually determined)


```{r}
cutoffs <- tibble(Patient = unique(average.CITE$Patient),
                   CD39 = c(3, 2.5, 2.5, 3.5, 6.5, 5, 1.5, 150),
                   CD103 = c(2.5, 2, 1, 2.8, 2.5, 3, 0.5, 100))
```

Plot 

```{r fig.asp=.65, fig.width=6}
average.CITE %>% 
#  filter(Patient != "NSCLC_37") %>% 
  mutate(Patient = factor(Patient, levels = unique(Patient))) %>% 
  ggplot(aes(x = UMIA.CD39, y = UMIA.CD103))+
  geom_point(aes(size = n, fill = as.character(MetaCell)), shape = 21, color = "black")+
  facet_rep_wrap(~Patient, nrow = 2, repeat.tick.labels = T, scales = "free")+
  theme_classic()+
  scale_y_log10(expand = expansion(mult = c(0.1, .05)))+
  scale_x_log10(expand = expansion(mult = c(0.1, .05)))+
  geom_hline(data = cutoffs, aes(yintercept = CD103), linetype = "dotted")+
  geom_vline(data = cutoffs, aes(xintercept = CD39), linetype = "dotted")+
  theme(legend.position = "none")

ggsave(filename = "Figs/Populations_definition.pdf", width = 11, height = 6, useDingbats = F, scale = .5)
```



## Dysfunction/Cytox signature

Import dysfunction signature

```{r }
imm.genes <- read_table("Data/dysfunction_score.txt") %>% pull(Gene)
```

Get expression of these genes.

```{r}
expression.data.dys <- get_expression(imm.genes)
```


Calculate dysfunction score from signature

```{r}

mc@mc %>% 
  enframe(name = "cellcode", value = "MetaCell") %>% 
  inner_join(patients) %>% 
  inner_join(expression.data.dys) %>% 
  mutate(MetaCell = factor(MetaCell, levels = as.character(1:85))) %>% 
  pivot_longer(cols = any_of(imm.genes), names_to = "Gene", values_to = "count") %>% 
  group_by(MetaCell, Patient) %>% 
  summarise(dys.score = (sum(count)/n())*1000) -> dysfunction.scores
```

Get cytox signature

```{r get_expression}
imm.genes <- read_table("Data/cytotox_score.txt") %>% pull(Gene)
```


Get expression of these genes.

```{r}
expression.data.cytox <- get_expression(imm.genes)
```

Calculate Cytox score


```{r}
mc@mc %>% 
  enframe(name = "cellcode", value = "MetaCell") %>% 
  inner_join(patients) %>% 
  inner_join(expression.data.cytox) %>% 
  mutate(MetaCell = factor(MetaCell, levels = as.character(1:85))) %>% 
  pivot_longer(cols = any_of(imm.genes), names_to = "Gene", values_to = "count") %>% 
  group_by(MetaCell, Patient) %>% 
  summarise(cytox.score = (sum(count)/n())*1000) -> cytox.scores
```


## correlation with CITE-seq

combine with CITE-seq data and calculate correlations

```{r}
average.CITE %>% 
  filter(Patient != "NSCLC_37") %>% 
  ungroup() %>% 
  inner_join(dysfunction.scores) %>% 
  inner_join(cytox.scores) %>% 
  select(-UMIA.CD4.1) %>% 
  filter(n > 5) %>% 
  group_by(Patient) %>% 
#  count()
  cor_test(vars = contains("UMI"), vars2 = contains("score"), method = "spearman") %>% 
  adjust_pvalue() -> cor.stats

write_tsv(cor.stats, "Output/scRNAseq_Correlations_Scores_ADT.tsv")

cor.stats
```


Plot

```{r}
cor.stats %>% 
  mutate(var1 = fct_reorder(var1, p, median)) %>% 
  ggplot(aes(x = var1, y = cor))+
  geom_boxplot()+
  geom_point(shape = 21, position = position_jitter(width = .1), aes(fill = var2))+
  facet_rep_wrap(~var2)+
  scale_fill_manual(values = c("green4", "red4"))+
  geom_hline(yintercept = 0)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        panel.grid.major.y = element_line(color = "grey90"), legend.position = "bottom")

ggsave("Figs/scRNAseq_scores_correlation_boxes.pdf", width = 4, height = 3)
```




```{r fig.asp=.45, fig.width=9}
average.CITE %>% 
  inner_join(dysfunction.scores) %>% 
  mutate(dys.score = case_when(dys.score > 600 ~ 600, TRUE ~ dys.score)) %>% 
  filter(Patient != "NSCLC_37") %>% 
  filter(n > 5) %>% 
  mutate(Patient = factor(Patient, levels = unique(Patient))) %>% 
  arrange(dys.score) %>% 
  ggplot(aes(x = UMIA.CD39, y = UMIA.CD103))+
  geom_point(aes(size = n, fill = (dys.score)), shape = 21, color = "black")+
  facet_rep_wrap(~Patient, nrow = 2, repeat.tick.labels = T, scales = "free")+
  theme_classic()+
  scale_y_log10(expand = expansion(mult = c(0.1, .05)))+
  scale_x_log10(expand = expansion(mult = c(0.1, .05)))+
  geom_hline(data = cutoffs, aes(yintercept = CD103), linetype = "dotted")+
  geom_vline(data = cutoffs, aes(xintercept = CD39), linetype = "dotted")+
 # scale_fill_gradient(low = "white", high = "red4")
  scale_fill_viridis_c(option = "plasma")

ggsave(filename = "Figs/scRNAseq_Populations_definition_dys.pdf", width = 11, height = 6, useDingbats = F, scale = .5)
```

```{r fig.asp=.7, fig.width=4}
average.CITE %>% 
  inner_join(dysfunction.scores) %>%
  inner_join(cytox.scores) %>% 
  filter(Patient != "NSCLC_37") %>% 
  group_by(MetaCell) %>% 
  summarise(n = sum(n), UMIA.CD39 = median(UMIA.CD39), UMIA.CD103 = median(UMIA.CD103), 
            dys.score = median(dys.score), cytox.score = median(cytox.score)) %>% 
  mutate(dys.score = case_when(dys.score > 600 ~ 600, TRUE ~ dys.score)) %>% 
  arrange(n) -> for.plot 

write_tsv(for.plot, "Output/scRNAseq_DysCytox_scores_CD39CD103.tsv")

```


Plot summarised scores


```{r fig.width=4}
p1 <- ggplot(for.plot, aes(x = UMIA.CD39, y = dys.score))+
  geom_point(size = 2, shape = 21, fill = "red4")+
  theme_classic()+
  scale_x_log10(expand = expansion(mult = c(0.1, .05)))+
  theme(panel.grid.major = element_line(color = "grey90"))

p2 <- ggplot(for.plot, aes(x = UMIA.CD103, y = dys.score))+
  geom_point(size = 2, shape = 21, fill = "red4")+
  theme_classic()+
  scale_x_log10(expand = expansion(mult = c(0.1, .05)))+
  theme(panel.grid.major = element_line(color = "grey90"))

p3 <- ggplot(for.plot, aes(x = UMIA.CD39, y = cytox.score))+
  geom_point(size = 2, shape = 21, fill = "green4")+
  theme_classic()+
  scale_x_log10(expand = expansion(mult = c(0.1, .05)))+
  theme(panel.grid.major = element_line(color = "grey90"))

p4 <- ggplot(for.plot, aes(x = UMIA.CD103, y = cytox.score))+
  geom_point(size = 2, shape = 21, fill = "green4")+
  theme_classic()+
  scale_x_log10(expand = expansion(mult = c(0.1, .05)))+
  theme(panel.grid.major = element_line(color = "grey90"))
```



```{r fig.width=5}
ggarrange(plotlist = list(p1,p2,p3,p4), common.legend = T, legend = "bottom" )

ggsave("Figs/scRNAseq_scores_makers_by_scores.pdf", height = 3.7, width = 4)
```




```{r}
ggplot(for.plot, aes(x = UMIA.CD39, y = UMIA.CD103))+
  geom_point(aes(fill = dys.score), size = 2.5, shape = 21, color = "black")+
  theme_classic()+
  scale_x_log10(expand = expansion(mult = c(0.1, .05)))+
  scale_y_log10(expand = expansion(mult = c(0.1, .05)))+
  scale_fill_viridis_c(option = "plasma")+
  geom_hline(yintercept = 1.0, linetype = "dotted")+
  geom_vline(xintercept = 3, linetype = "dotted")+
  theme(panel.grid.major = element_line(color = "grey90"))
  
ggsave("Figs/scRNAseq_scores_makers_by_dys_simple.pdf", height = 2, width = 3.1)
```




