---
title: "scRNAseq NSCLC population analyses"
author: "Kaspar Bresser"
date: "18/12/2025"
output: 
  github_document:
    toc: true
  html_document: 
    theme: simplex
    highlight: pygments
#    code_folding: show
    self_contained: TRUE
    toc: yes
    toc_float:
      collapsed: no
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
                      fig.width = 5,
                      fig.asp = 0.618,
                      fig.align = "center")
```

Below some general visualizations of the scRNAseq data. 

```{r loading, echo=TRUE}
library(Seurat)
library(metacell)
library(tidyverse)
library(lemon)
library(ggrastr)
```


## MetaCell projection and composition

Import the MetaCell object and the 2D coordinates.

```{r import_mc}
scdb_init("Data")
mc <- scdb_mc("NSCLC_MC_reordered")

str(mc)
```


```{r import_coords}
MC.graph <- scdb_mc2d("NSCLC_mc2d_r")

coords <- tibble( cellcode = names(MC.graph@sc_y),
                  sc.x = MC.graph@sc_x,
                  sc.y = MC.graph@sc_y)
coords
```

## 2d projection

Add the MetaCell identities to the coordinate data, and plot the 2D projection.

Calculate centroids to plot MC ids on top. 

```{r}

df <- mc@mc %>% 
  enframe(name = "cellcode", value = "MetaCell") %>% 
  inner_join(coords) %>% 
  mutate(MetaCell = as.factor(MetaCell))

# count + centroid per metacell
centroids <- df %>%
  group_by(MetaCell) %>%
  summarise(
    n = n(),
    x = mean(sc.x),
    y = mean(sc.y),
    .groups = "drop"
  )
```


```{r}
ggplot(df, aes(x = sc.x, y = sc.y, color = MetaCell)) +
 rasterize( geom_point(size = 0.8), dpi = 600) +
  
  # circle
  geom_point(
    data = centroids,
    aes(x = x, y = y),
    shape = 21,
    size = 6,
    fill = "white",
    color = "black",
    stroke = 0.6
  ) +
  
  # number inside
  geom_text(
    data = centroids,
    aes(x = x, y = y, label = MetaCell),
    size = 2
  ) +
  
  theme_classic() +
  theme(legend.position = "none")

ggsave(filename = "Figs/scRNAseq_2d_proj_MCs_labeled.pdf", width = 6, height = 6)
```

Plot number of cells per MC

```{r}
df %>% 
  ggplot(aes(x = as.numeric(MetaCell), fill = MetaCell)) + 
  geom_bar() +
  scale_x_continuous(breaks = seq(0,85,10))+
  theme(legend.position = "none", panel.grid = element_line(color = "grey95"))

ggsave(filename = "Figs/scRNAseq_cellcount_per_MC.pdf", width = 6, height = 2.5)
```


ADT count per MC

```{r}
mat.obj <- scdb_mat("NSCLC_filt")

mat.obj@cell_metadata %>% 
  as_tibble(rownames = "cellcode") %>% 
  inner_join(df) %>% 
  group_by(MetaCell) %>% 
  summarise(nCount.reads = sum(nCount_RNA), nCount = sum(nCount_ADT)) %>% 
ggplot(aes(x = as.numeric(MetaCell), fill = MetaCell, y = nCount/1000)) + 
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = seq(0,85,10))+
  theme(legend.position = "none", panel.grid = element_line(color = "grey95"))

ggsave(filename = "Figs/scRNAseq_ADTcount_per_MC.pdf", width = 6, height = 2.5)
```




## Patient composition

Extract the patient info from the mat object

```{r get_hashtags}
mat.obj <- scdb_mat("NSCLC_filt")

mat.obj@cell_metadata %>% 
  select(Patient) %>% 
  as_tibble(rownames = "cellcode") -> patients

patients
```

Combine with MetaCells and normalize within samples. 

```{r calc_composition}
mc@mc %>% 
  enframe(name = "cellcode", value = "MetaCell") %>% 
  inner_join(patients) %>% 
  count(MetaCell, Patient) %>%
  group_by(Patient)%>%
  mutate(normalized.count = (n/sum(n))*1000 ) %>% 
  mutate(MetaCell = as.factor(MetaCell) ) -> sample.counts.MC

sample.counts.MC
```




## Define populations

I'll try to define the populations that I sorted on using the CITEseq hashtag information. 


Import the CITEseq information

```{r}
mat.obj@cell_metadata %>% 
  as_tibble(rownames = "cellcode") %>% 
  select(cellcode, contains("UMIA")) -> CITE

CITE
```

Take the median of each CITEseq antibody per MetaCell

```{r}
MC.ids <- enframe(mc@mc, name = "cellcode", value = "MetaCell") %>% mutate(MetaCell = as.character(MetaCell))

CITE %>% 
  inner_join(MC.ids) %>% 
  inner_join(patients) %>% 
  group_by(Patient) %>% 
  mutate(UMIA.CD39 = ((UMIA.CD39+1)/sum(UMIA.CD39))*10000,
         UMIA.CD103 = ((UMIA.CD103+1)/sum(UMIA.CD103))*10000) %>% 
  group_by(MetaCell, Patient) %>% 
  summarise(across(contains("UMIA"), median)) -> average.CITE

average.CITE
```



```{r}
sample.counts.MC %>% 
  mutate(MetaCell = (MetaCell)) %>% 
  inner_join(average.CITE) -> for.plot
```



Visualize per patient, and set some cutoffs


```{r}
cutoffs <- tibble(Patient = unique(for.plot$Patient),
                   CD39 = c(3, 2.5, 2.5, 3.5, 6.5, 5, 1.5, 150),
                   CD103 = c(2.5, 2, 1, 2.8, 2.5, 3, 0.5, 100))
```

Plot 

```{r fig.asp=.65, fig.width=6}
for.plot %>% 
#  filter(Patient != "NSCLC_37") %>% 
  mutate(Patient = factor(Patient, levels = unique(for.plot$Patient))) %>% 
  ggplot(aes(x = UMIA.CD39, y = UMIA.CD103))+
  geom_point(aes(size = normalized.count, fill = as.character(MetaCell)), shape = 21, color = "black")+
  facet_rep_wrap(~Patient, nrow = 2, repeat.tick.labels = T, scales = "free")+
  theme_classic()+
  scale_y_log10(expand = expansion(mult = c(0.1, .05)))+
  scale_x_log10(expand = expansion(mult = c(0.1, .05)))+
  geom_hline(data = cutoffs, aes(yintercept = CD103), linetype = "dotted")+
  geom_vline(data = cutoffs, aes(xintercept = CD39), linetype = "dotted")+
  theme(legend.position = "none")

ggsave(filename = "Figs/Populations_definition.pdf", width = 11, height = 6, useDingbats = F, scale = .5)
```


## Subset into populations

Based on the cutoffs, divide the cells from each MC to the DN, SP, DP populations. Do this for each patient.


```{r}
for.plot %>% 
  select(MetaCell, Patient, n, UMIA.CD39, UMIA.CD103) %>% 
  pivot_longer(cols = c(UMIA.CD39, UMIA.CD103)) %>% 
  mutate(name = str_remove(name, "UMIA\\.")) %>% 
  inner_join(pivot_longer(cutoffs, cols = c(CD39, CD103), names_to = "name", values_to = "cutoff")) %>% 
  mutate(tmp = value > cutoff) %>% 
  group_by(MetaCell, Patient, n) %>% 
  summarise(population = sum(tmp)) %>%
  mutate(population = case_when(population == 0 ~ "DN",
                                population == 1 ~ "SP",
                                population == 2 ~ "DP")) -> MC.populations
```

If a MC could be in multiple populations across the patients, assign it to the dominant population.

```{r}
MC.populations %>% 
  group_by(MetaCell, population) %>% 
  summarise(n = sum(n)) %>% 
  group_by(MetaCell) %>% 
  reframe(n.prop = n/sum(n), population = population) %>% 
  group_by(MetaCell) %>% 
  slice_max(n.prop, n=1) %>% 
  select(MetaCell, population) -> population.id
```


## Gene expression analyses 

Import Seurat object and Normalize the data


```{r}
seurat.obj <- read_rds("Data/CD8_NSCLC_filt.rds")
seurat.obj <- NormalizeData(seurat.obj, assay = "RNA", normalization.method = "CLR")
```

Define function to extract counts per gene per cellcode from the seurat object


```{r}
get_expression <- function(gene.list){
  
 seurat.obj %>% 
  GetAssayData( layer = "data", assay = "RNA")%>% 
  t() %>% 
  as.data.frame() %>% 
  as_tibble(rownames = "cellcode") %>% 
  select(one_of(c("cellcode", gene.list)))
  
}
```




Define Immune related genes

```{r get_expression}
imm.genes <- c('ITGAE', 'ENTPD1','CXCL13', "CDK4", 'LEF1', 'GZMA', 'TOX','PDCD1', 'HAVCR2', 'TCF7', 'GZMK', "GIMAP4" )
```

Get expression of these genes.

```{r}
expression.data <- get_expression(imm.genes)
```

Plot their expression across the populations. Dots are MCs in this one.

```{r fig.width=7}
expression.data %>% 
  inner_join(MC.ids) %>% 
  inner_join(population.id) %>% 
  pivot_longer(cols = all_of(imm.genes), names_to = "gene", values_to = "norm_umi") %>% 
  group_by(gene, MetaCell, population) %>% 
  summarise(expr = mean(norm_umi)) %>% 
  mutate(population = factor(population, levels = c("DN", "SP", "DP"))) %>% 
ggplot(aes(x= population, y = expr, fill = population))+
  geom_boxplot(outlier.shape = NA, color ="black")+
  geom_jitter(width = 0.15, size = 0.33)+
  facet_rep_wrap(~gene, scales = "free")+
  scale_fill_manual(values = c("grey", "#fc9272","#cb181d"))+
  theme(panel.grid.major.y = element_line(color = "grey90"))

ggsave(filename = "Figs/scRNAseq_Populations_ImmGene_perMC.pdf", width = 6, height = 4, useDingbats=FALSE)
```



Plot their expression across the populations. Dots are Patients in this one.

```{r fig.width=7}
expression.data %>% 
  inner_join(MC.ids) %>% 
  inner_join(population.id) %>% 
  inner_join(patients) %>% 
  pivot_longer(cols = all_of(imm.genes), names_to = "gene", values_to = "norm_umi") %>% 
  group_by(gene, Patient, population) %>% 
  summarise(expr = mean(norm_umi)) %>% 
  mutate(population = factor(population, levels = c("DN", "SP", "DP"))) %>% 
ggplot(aes(x= population, y = expr, fill = population))+
  geom_boxplot(outlier.shape = NA, color ="black")+
  geom_jitter(width = 0.15, size = 1.75, shape = 21)+
  facet_rep_wrap(~gene, scales = "free")+
  scale_fill_manual(values = c("grey", "#fc9272","#cb181d"))+
  theme(panel.grid.major.y = element_line(color = "grey90"))

ggsave(filename = "Figs/scRNAseq_Populations_ImmGene_perPatient.pdf", width = 6, height = 4, useDingbats=FALSE)
```


## Export files

Export population identity of MetaCells 

```{r}
expression.data %>% 
  inner_join(MC.ids) %>% 
  inner_join(population.id) %>% 
  inner_join(patients) %>% 
  select(-c(imm.genes)) %>% 
  write_tsv("Output/scRNAseq_population_per_cellcode.tsv")
```


Export pseudobulked data across populations

```{r}
seurat.obj %>% 
  GetAssayData( layer = "counts", assay = "RNA")%>% 
  t() %>% 
  as.data.frame() %>% 
  as_tibble(rownames = "cellcode") %>% 
  inner_join(MC.ids) %>% 
  inner_join(population.id) %>% 
  inner_join(patients) -> all.expr.data
```

```{r}
all.expr.data %>% 
  pivot_longer(cols = all_of(seurat.obj@assays[["RNA"]]@counts@Dimnames[[1]]), names_to = "gene", values_to = "count") %>% 
  group_by(gene, Patient, population) %>% 
  summarise(count = sum(count)) -> all.expr.data
```


```{r}
write_tsv(all.expr.data, "Output/scRNAseq_pseudobulked_readcounts.tsv")
```

