---
title: "DE testing"
author: "Kaspar Bresser"
output: 
  github_document:
    toc: true
 # html_document: 
#    theme: simplex
#    highlight: pygments
#    code_folding: show
#    self_contained: TRUE
#    toc: yes
#    toc_float:
#      collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
                      fig.width = 7,
                      fig.asp = 0.618,
                      fig.align = "center")
```


```{r loading, echo=FALSE}
library(rstatix)
library(ggpubr)
library(gghighlight)
library(lemon)
library(RColorBrewer)
library(edgeR)
library(tidyverse)
library(clusterProfiler)
library(org.Hs.eg.db)

```


## Import and inspect

Import reads

```{r filter_aligned}
read.counts <- read_tsv( "Output/scRNAseq_pseudobulked_readcounts.tsv") %>% mutate(Patient = str_remove(Patient, "_"))
```

pivot to wide data

```{r}
read.counts %>% 
 # filter(population != "DN") %>% 
  filter(Patient != "NSCLC37") %>% 
  pivot_wider(names_from = c(Patient, population), values_from = count) -> read.counts.wide
  
```


Generate DGE object

```{r}
DGE.object <- DGEList(counts = read.counts.wide[2:22], 
                      genes = pull(read.counts.wide, gene))
```


Filter low expression

```{r}
countsPerMillion <- as.data.frame(cpm(DGE.object), row.names =  pull(read.counts.wide, gene))

thresh <- countsPerMillion > 4
table(rowSums(thresh))

summary(rowSums(thresh) >= 5)
```

extract CPM

```{r}
DGE.object <- DGE.object[rowSums(thresh) >= 5, , keep.lib.sizes=FALSE]
DGE.object <- calcNormFactors(DGE.object)

countsPerMillion <- as.data.frame(cpm(DGE.object, normalized.lib.sizes = T, log = T), 
                                  row.names = DGE.object$genes$genes)
```


## PCA

get variance

```{r}
var_genes <- apply(countsPerMillion, 1, var)
head(var_genes)
```

select top 1,000, perform PCA

```{r}
select_var <- names(sort(var_genes, decreasing=TRUE))[1:1000]
countsPerMillion <- countsPerMillion[select_var, ]

countsPerMillion %>% 
  t() %>% 
  as.data.frame() %>% 
  prcomp(center = T, scale = F) -> pca

str(pca)
```

Plot PCA

```{r fig.width=4, fig.asp=.8}
pca %>% 
  augment() %>% 
  separate(.rownames, into = c("Patient", "population")) %>% 
  mutate(population = factor(population, levels = c("DN", "SP", "DP"))) %>% 
ggplot(aes(x = .fittedPC1, y = .fittedPC2))+
  geom_point( size = 3, shape = 21, aes( fill = population))+
     stat_ellipse(geom = "polygon",
               aes(fill = paste( population)), 
               alpha = .2)+
  scale_fill_manual(values = c("grey", "#fc9272","#cb181d"))+
  ggtitle("PCA 1000 most variable genes")+
  theme_classic()+
  theme(panel.grid.major = element_line(color = "grey90"))+
  guides(fill = guide_legend("timepoint", override.aes = list(shape = 21)))
```


Plot heatmap

```{r}
select_var <- names(sort(var_genes, decreasing=TRUE))[1:1000]
countsPerMillion <- countsPerMillion[select_var,]

as.data.frame(pca$x) %>% 
  rownames_to_column("sample") %>% 
  separate(sample, into = c("donor", "population"), remove = F) %>% 
  dplyr::select(sample, donor,  population) %>% 
#  filter(population != "GFPpos") %>% 
  column_to_rownames("sample") -> annotation

annotation
```


```{r fig.asp=.8, fig.width=9}
library(pheatmap)

countsPerMillion %>% 
#  select(!contains("GFPpos")) %>% 
pheatmap(scale = "row",breaks = seq(-2,2,by=0.04), 
         clustering_distance_cols = "correlation" , 
         annotation_col = annotation, show_rownames = F,
         border_color = NA, clustering_distance_rows = "euclidean", 
         main = "1000 most variable genes", cutree_cols = 2, cutree_rows = 2)
```





## DE analysis - edgeR

Set up design

```{r}
DGE.object$group <- rep(c("DN", "DP", "SP"), 7)
g <- factor((DGE.object$group))


# Set design for comparison
design <- model.matrix(~ 0 + g)
colnames(design) <- levels(g)

design

# Estimate dispersions
DGE.object <- estimateDisp(DGE.object, design, robust=TRUE)

plotMDS(DGE.object)
```


estimate fit, and test for DP versus SP contrast

```{r}
fit <- glmQLFit(DGE.object, design)

my.contrasts <- makeContrasts(DPvSP = DP-SP, levels=design)


qlf <- glmQLFTest(fit, contrast = my.contrasts[,"DPvSP"])

plotMD(qlf)
```

extract and output results


```{r}
qlf$table %>% 
  as_tibble() %>% 
  mutate(gene = DGE.object$genes$genes) %>% 
  mutate(FDR = p.adjust(PValue, method = "fdr")) -> DE.results

DE.results

write_tsv(DE.results, "Output/scRNAseq_DE_results.tsv")
```


## GSEA 

Set up gene list

```{r}
gene.list <- DE.results %>%
  arrange(desc(logFC)) %>%
  mutate(ENTREZID = mapIds(org.Hs.eg.db,
                           keys = gene,
                           keytype = "SYMBOL",
                           column = "ENTREZID")) %>%
  filter(!is.na(ENTREZID)) %>%
  dplyr::select(ENTREZID, logFC)

# Named numeric vector for GSEA
gene.list.vec <- gene.list$logFC
names(gene.list.vec) <- gene.list$ENTREZID
```

run GSEA on Biological Process (GO)

```{r}
gsea.res <- gseGO(
  geneList = sort(gene.list.vec, decreasing = TRUE),
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  keyType = "ENTREZID",
  pvalueCutoff = 0.05
)

dotplot(gsea.res, showCategory = 15)
```

Check ImmuneSigDB


```{r fig.asp=1}
library(msigdbr)

# Get human immunologic signatures
m.t2 <- msigdbr(species = "Homo sapiens", collection = "C7")

# Convert to list for clusterProfiler
m.t2 <-  m.t2[, c("gs_name", "gene_symbol")]


# Keep only pathways containing "TCELL"
m.t2 <- m.t2 %>%
  filter(grepl("TCELL", gs_name, ignore.case = TRUE))


gene.list.vec <- DE.results$logFC
names(gene.list.vec) <- DE.results$gene
gene.list.vec <- sort(gene.list.vec, decreasing = TRUE)

# Run GSEA
gsea.res <- GSEA(
  geneList = gene.list.vec, # logFC ranked
  TERM2GENE = m.t2,
  pvalueCutoff = 0.05
)
```


Then plot individual pathways

```{r fig.width=4, fig.asp=.7}
library(enrichplot)
gseaplot2(gsea.res, geneSetID = c("GSE24081_CONTROLLER_VS_PROGRESSOR_HIV_SPECIFIC_CD8_TCELL_DN",
                                  "GSE24081_CONTROLLER_VS_PROGRESSOR_HIV_SPECIFIC_CD8_TCELL_UP"), 
          title = "CONTROLLER_VS_PROGRESSOR_HIV_SPECIFIC_CD8_TCEL", pvalue_table = F,
          subplots = 1)+
  geom_hline(yintercept = 0, linetype = "solid")+
  theme(legend.position = "bottom")

ggsave("Figs/scRNAseq_DPvsSP_HIV.pdf", width = 5, height = 4)
```


```{r fig.width=4, fig.asp=.7}
gseaplot2(gsea.res, geneSetID = c("GSE30962_ACUTE_VS_CHRONIC_LCMV_PRIMARY_INF_CD8_TCELL_DN",
                                  "GSE30962_ACUTE_VS_CHRONIC_LCMV_PRIMARY_INF_CD8_TCELL_UP"), 
          title = "ACUTE_VS_CHRONIC_LCMV_PRIMARY_INF_CD8_TCELL", pvalue_table = F,
          subplots = c(1))+
  geom_hline(yintercept = 0, linetype = "solid")+
  theme(legend.position = "bottom", panel.grid.major.y = element_line())

ggsave("Figs/scRNAseq_DPvsSP_LCMV.pdf", width = 5, height = 4)
```


## plot genes

```{r}
dat.cpm <- as.data.frame(cpm(DGE.object, normalized.lib.sizes = T, log = T), 
                                  row.names = DGE.object$genes$genes) %>% as_tibble(rownames = "gene")
```


```{r fig.asp=.9}
imm.genes <- c('ITGAE', 'ENTPD1',  'HAVCR2', 'PDCD1', "TIGIT", 'TOX',
               "MKI67", "CDK4",'CXCL13', "IL17A", "IL2", "TNF",
                "LEF1", 'TCF7','GZMK', "IL7R", "CCR7", "JUNB" )

dat.cpm %>% 
  filter(gene %in% imm.genes) %>% 
  pivot_longer(-gene, names_to = "sample", values_to = "CPM") %>% 
  separate(sample, into = c("Patient", "population")) %>% 
  mutate(population = factor(population, levels = c("DN", "SP", "DP"))) %>% 
  mutate(gene = factor(gene, levels = imm.genes)) %>% 
ggplot(aes(x= population, y = CPM, fill = population))+
  geom_boxplot(outlier.shape = NA, color ="black")+
  geom_jitter(width = 0.15, size = 1.75, shape = 21)+
  facet_rep_wrap(~gene, scales = "free", nrow =3)+
  scale_fill_manual(values = c("grey", "#fc9272","#cb181d"))+
  theme_classic()+
  theme(panel.grid.major.y = element_line(color = "grey90"))

ggsave("Figs/scRNAseq_examples_pseudobulk.pdf", width = 7, height = 5)
```





